Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal

### 2020-08-25 Manuscript/SupportingInfo.Rmd
SI

---
title: "Supporting Information for: A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal"

author: "Dean C. Adams, Erica K. Baken, and Michael L. Collyer"

@@

Here we provide additional supporting information referenced in the main document, which include additional analyses, and simulation results across a wider set of input conditions. As before, simulations were conducted on six different tree sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylogenetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). We generated 100 random trees for each intersection of tree size and $\lambda$. For each $\lambda$ within each tree size, continuous traits were then simulated on each phylogeny under a BM model of evolution. For each set of 100 trees we measured the mean values of $\hat{\lambda}$ and $\kappa$, their standard deviation, and calculated the Shapiro-Wilk $W$ statistic as a departure from normality (symmetry). For the latter, a value of $1.0$ indicates normally distributed values, while departures from $1.0$ indicate skewness.

@@

Results from the simulations on pure-birth trees revealed that both $Z_{\lambda}$ and $Z_K$ increased with increasing input phylogenetic signal, but $\hat{\lambda}$ was strongly affected by tree size, whereas $Z_K$ was m1re consistent (Fig. S4). Additionally, $Z_K$ increased more linearly with increasing levels of phylogenetic signal, and its standard deviation across input signal was more even across tree sizes, implying more consistent precision. This latter observation is seen in the upper four panels of Fig. S1, where the dispersion of simulation results (gray dots) is far more even at each input level of phylogenetic signal for $Z_K$ than it is for $Z_{\lambda}$.

@@

For *K*, mean values increased with increasing phylogenetic signal, though as was found with pure-birth trees, the increase was nonlinear (Fig. S3 black line). Likewise, variation increased with increasing phylogenetic signal (Fig. S3 red line), though the distribution of *K* was more normally distributed throughout its range, and across different tree sizes, as compared with $\hat{\lambda}$, thought there was some slight skew for high input levels of phylogenetic signal on large phylogenies (Fig. S3 blue line). \hfill\break

@@

We also investigaged the effect of unresolved phylogenies on estimates of phylogenetic signal by adjusting pure-birth trees to have 50% collapsed nodes, following the procedures of [@MolinaVenegas2017]. Results from simulations on phylogenies containing polytomies mirrored those found on fully resolved trees. For $\hat{\lambda}$, the mean value increased with increasing input signal, but was negatively biased, and was less than the input value across most of its range (Fig. S8: black line). Additionally, the precision of $\hat{\lambda}$ varied with differing input levels, with the greatest variation found at intermediate values of $\lambda$ (Fig. S8 red line). Finally, the distribution of $\hat{\lambda}$ was not normal, and became more skewed at more extreme values of $\lambda$ (Fig. S8 blue line).  \hfill\break

@@

We also investigated the effect of incorporating $\lambda$ in PGLS analyses (phylogenetic regression and ANOVA). Here, patterns for $\hat\lambda$, *K*, $Z_\lambda$ and $Z_\kappa$ were virtually identical to those found previously. In essence, $\hat\lambda$ was not suitable as a test statistic representing phylogenetic signal due to all properties previously shown (Figs. S11 & S12), *K* was more appropriate, but  associated non-linearly with signal strength (Figs. S13 & 14), while $Z_\kappa$ was found to be a superior effect size as compared with $Z_\lambda$ (Figs S15 & S16). \hfill\break

@@

\textbf{Figure S11}. Response of Pagel's \(\lambda\) to increasing strength of Brownian motion in phylogenetic regression (incorporating $\lambda$). Gray line signifies the 1:1 line where the input value matches the estimate.

@@

\textbf{Figure S12}. Response of Pagel's \(\lambda\) to increasing strength of Brownian motion in phylogenetic ANOVA (incorporating $\lambda$). Gray line signifies the 1:1 line where the input value matches the estimate.

@@

\textbf{Figure S13}. Response of Blomberg's \textit{K} to increasing strength of Brownian motion in phylogenetic regression (incorporating $\lambda$). Gray line signifies the 1:1 line where the input value matches the estimate.

@@

\textbf{Figure S14}. Response of Blomberg's \textit{K} to increasing strength of Brownian motion in phylogenetic ANOVA (incorporating $\lambda$). Gray line signifies the 1:1 line where the input value matches the estimate.

@@

\textbf{Figure S17}. Parameter estimates from phylogenetic regression (incorporating $\lambda$). Gray line is the 1:1 line and red line is the estimate.

@@

\textbf{Figure S18}. Parameter estimates from phylogenetic ANOVA (incorporating $\lambda$). Gray line is the 1:1 line and red line is the estimate.

@@

\textbf{Figure S19}. Power curves for phylogenetic regression (incorporating $\lambda$). Type I error is found on the far-left of the plot.

@@

\textbf{Figure S20}. Power curves for phylogenetic ANOVA (incorporating $\lambda$). Type I error is found on the far-left of the plot.

>>>>

---
title:

author:

@@

```{r, echo = FALSE, out.width="95%", align= "center"}
include_graphics("SI-Title.png")  
```
\hfill\break
\hfill\break

## Supplementary Information for 
\hfill\break

A Standardized Effect Size for Evaluating and Comparing the Strength of Phylogenetic Signal \hfill\break
\hfill\break
Dean C. Adams, Erica K. Baken, and Michael L. Collyer \hfill\break

\hfill\break
Dean C. Adams\hfill\break
Email:  dcadams@iastate.edu\hfill\break

## This PDF file includes:
\hfill\break
- Supplementary text 
- Figures S1 to S20 
- SI References

Here we provide additional supporting information referenced in the main document, which include additional analyses, and simulation results across a wider set of input conditions. As before, simulations were conducted on six different tree sizes ($n=2^5, 2^6, \cdots, 2^{10}$), and with differing levels of phylogenetic signal ($\lambda=0.0, 0.5, \cdots, 1.0$). We generated 50 random trees for each intersection of tree size and $\lambda$. For each $\lambda$ within each tree size, continuous traits were then simulated on each phylogeny under a BM model of evolution. For each set of 50 trees we measured the mean values of $\hat{\lambda}$ and $K$, their standard deviation, and calculated the Shapiro-Wilk $W$ statistic as a departure from normality (symmetry). For the latter, a value of $1.0$ indicates normally distributed values, while departures from $1.0$ indicate skewness.

@@

Results from the simulations on pure-birth trees revealed that both $Z_{\lambda}$ and $Z_K$ increased with increasing input phylogenetic signal, but $\hat{\lambda}$ was strongly affected by tree size, whereas $Z_K$ was more consistent (Fig. S4). Additionally, $Z_K$ increased more linearly with increasing levels of phylogenetic signal, and its standard deviation across input signal was more even across tree sizes, implying more consistent precision. This latter observation is seen in the upper four panels of Fig. S1, where the dispersion of simulation results (gray dots) is far more even at each input level of phylogenetic signal for $Z_K$ than it is for $Z_{\lambda}$.

@@

For *K*, mean values increased with increasing phylogenetic signal, though as was found with pure-birth trees, the increase was nonlinear (Fig. S3 black line). Likewise, variation increased with increasing phylogenetic signal (Fig. S3 red line), though the distribution of *K* was more normally distributed throughout its range, and across different tree sizes, as compared with $\hat{\lambda}$, though there was some slight skew for high input levels of phylogenetic signal on large phylogenies (Fig. S3 blue line). \hfill\break

@@

We also investigaged the effect of unresolved phylogenies on estimates of phylogenetic signal by adjusting pure-birth trees to have 20% collapsed nodes, following the procedures of [@MolinaVenegas2017]. Results from simulations on phylogenies containing polytomies mirrored those found on fully resolved trees. For $\hat{\lambda}$, the mean value increased with increasing input signal, but was negatively biased, and was less than the input value across most of its range (Fig. S8: black line). Additionally, the precision of $\hat{\lambda}$ varied with differing input levels, with the greatest variation found at intermediate values of $\lambda$ (Fig. S8 red line). Finally, the distribution of $\hat{\lambda}$ was not normal, and became more skewed at more extreme values of $\lambda$ (Fig. S8 blue line).  \hfill\break

@@

We also investigated the effect of incorporating $\lambda$ in PGLS analyses (phylogenetic regression and ANOVA). Here, patterns for $\hat\lambda$, *K*, $Z_\lambda$ and $Z_K$ were virtually identical to those found previously. In essence, $\hat\lambda$ was not suitable as a test statistic representing phylogenetic signal due to all properties previously shown (Figs. S11 & S12), *K* was more appropriate, but  associated non-linearly with signal strength (Figs. S13 & 14), while $Z_K$ was found to be a superior effect size as compared with $Z_\lambda$ (Figs S15 & S16). \hfill\break

@@

\textbf{Figure S11}. Response of Pagel's \(\lambda\) to increasing strength of Brownian motion in phylogenetic regression (incorporating $\lambda$). Gray line signifies the 1:1 line where the input value matches the estimate. At each input level, the dark black line represents the empirically derived expected value (mean) of \(\hat\lambda\), the red line is the standard deviation of \(\hat\lambda\), and the blue line is Shapiro Wilks statistic of \(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

@@

\textbf{Figure S12}. Response of Pagel's \(\lambda\) to increasing strength of Brownian motion in phylogenetic ANOVA (incorporating $\lambda$). Gray line signifies the 1:1 line where the input value matches the estimate. At each input level, the dark black line represents the empirically derived expected value (mean) of \(\hat\lambda\), the red line is the standard deviation of \(\hat\lambda\), and the blue line is Shapiro Wilks statistic of \(\hat\lambda\) (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

@@

\textbf{Figure S13}. Response of Blomberg's \textit{K} to increasing strength of Brownian motion in phylogenetic regression (incorporating $\lambda$).  At each input level, the dark black line represents the empirically derived expected value (mean) of \textit{K}, the red line is the standard deviation of \textit{K}, and the blue line is Shapiro Wilks statistic of \textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

@@

\textbf{Figure S14}. Response of Blomberg's \textit{K} to increasing strength of Brownian motion in phylogenetic ANOVA (incorporating $\lambda$). At each input level, the dark black line represents the empirically derived expected value (mean) of \textit{K}, the red line is the standard deviation of \textit{K}, and the blue line is Shapiro Wilks statistic of \textit{K} (\(W=1.0\) signifies normality, \(W< 1.0\) represent skewed distributions).

@@

\textbf{Figure S17}. Parameter estimates from phylogenetic regression (incorporating $\lambda$). The red line represents the estimated slope, which almost completely overlaps the gray 1:1 line.

@@

\textbf{Figure S18}. Parameter estimates from phylogenetic ANOVA (incorporating $\lambda$). The red line represents the estimated slope, which almost completely overlaps the gray 1:1 line.

@@

\textbf{Figure S19}. Power curves for phylogenetic regression (incorporating $\lambda$). Type I error is found on the far-left of the plot (effect = 0).

@@

\textbf{Figure S20}. Power curves for phylogenetic ANOVA (incorporating $\lambda$). Type I error is found on the far-left of the plot (effect = 0).
